---
title: "Explore UK Biobank data"
author: "Ken B. Hanscombe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<br>

After downloading and decrypting your UK Biobank (UKB) data with the supplied UKB programs (http://biobank.ctsu.ox.ac.uk/crystal/docs/UsingUKBData.pdf), you have multiple files that need to be brought together to give you a tidy dataset to explore. The data file has column names that are the field codes from the UKB data showcase. `ukbtools` provides tools to tidy up the naming of the variables and perform exploratory analysis. 

<br>
<br>


## 1. Preparing a UKB fileset

Download^ยง^ then decrypt your data and create a UKB fileset (.tab, .r, .html):

```{bash, eval = FALSE}
ukb_unpack ukbxxxx.enc key
ukb_conv ukbxxxx.enc_ukb r
ukb_conv ukbxxxx.enc_ukb docs

```


`ukb_unpack` decrypts your downloaded `ukbxxxx.enc` file, outputting a `ukbxxxx.enc_ukb` file. `ukb_conv` with the `r` flag converts the decrypted data to a tab-delimited file `ukbxxxx.tab` and an R script `ukbxxxx.r` that reads the tab file. The `docs` flag creates an html file containing a field-code-to-description table (among others).

<br>

^ยง^ Full details of the data download and decrypt process are given in the _Using UK Biobank Data_ documentation (http://biobank.ctsu.ox.ac.uk/crystal/docs/UsingUKBData.pdf).

<br>
<br>


## 2. Making a dataset

The function `ukb_df()` takes two arguments, the stem of your fileset and the path, and returns a dataframe with usable column names. 


```{r, eval = FALSE}

my_ukb_data <- ukb_df("ukbxxxx")

```


You can also specify the path to your fileset if it is not in the current directory. For example, if your fileset is in a subdirectory of the working directory called data


```{r, eval = FALSE}

my_ukb_data <- ukb_df("ukbxxxx", path = "./data/")

```


<br>

__Note:__ You can move the three files in your fileset after creating them with `ukb_conv`, but they should be kept together. `ukb_df()` automatically updates the read call in the R source file to point to the correct directory (the current directly by default, or the directory specified by `path`).

<br>
<br>

#### 2.1 Multiple downloads

If you have multiple UKB downloads, tidy then merge them.

```{r eval = FALSE}

ukbxxxx_data <- ukb_df("ukbxxxx", path = "./data/")
ukbyyyy_data <- ukb_df("ukbyyyy", path = "./data/")
ukbzzzz_data <- ukb_df("ukbzzzz", path = "./data/")

my_ukb_data <- plyr::join_all(
  list(
    ukbxxxx_data,
    ukbyyyy_data,
    ukbzzzz_data),
  by = "eid",
  type = "full")

```

<br>

__Note:__ You will have at least one variable duplicated in the different downloads, the unique individual identifier `eid`. Counting the repeated variables (and the number of repeats of each) can give you a sense of the expected number of variables in your merged dataset.

```{r, eval = FALSE}

ukb_names <- c(
  names(ukbxxxx_data),
  names(ukbyyyy_data),
  names(ukbzzzz_data)
)
count_ukb_names <- data.frame(table(ukb_names))
dim(count_ukb_names[count_names$Freq > 1, ])

```

<br>
<br>


## 3. ICD diagnoses




## 4. Genetic metadata




